---
title: "meddebt data manipulation"
author: "Rachel Marconi"
date: "2024-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(janitor)
library(dplyr)
library(readxl)
library(jsonlite)
```

# Import and join all data for main chart

## County level 
```{r}
# county share and amount data for all debtors
county_share_amount_all <- read_xlsx('./data-in/all_county_export.xlsx') %>% 
  clean_names()

# county share and amount data for POC and white breakdown
county_share_amount_race <- read_xlsx('./data-in/all_county_export_race.xlsx') %>% 
  clean_names() %>% 
  select(-state_abreviation)

county_share_amount <- county_share_amount_all %>% 
  left_join(county_share_amount_race, by = c("county_nm","county_fips","year")) %>% 
  rename(name = county_nm, fips = county_fips)

```

## State level
```{r}
# state share and amount data for all debtors
state_share_amount_all <- read_xlsx('./data-in/all_state_export.xlsx') %>% 
  clean_names()

# state share and amount data for POC and white breakdown
state_share_amount_race <- read_xlsx('./data-in/all_state_export_race.xlsx') %>% 
  clean_names()

state_share_amount <- state_share_amount_all %>% 
  left_join(state_share_amount_race, by = c("state_abreviation","state_fip","year")) %>% 
  rename(name = state_abreviation, fips = state_fip)

```

## National level
```{r}
# US share and amount data for all debtors
nation_share_amount_all <- read_xlsx('./data-in/all_usa_export.xlsx') %>% 
  clean_names() %>% 
  select(-usa)

nation_share_amount_race <- read_xlsx('./data-in/all_usa_export_race.xlsx') %>% 
  clean_names() %>% 
  select(-usa)

nation_share_amount <- nation_share_amount_all %>% 
  left_join(nation_share_amount_race, by = c("year")) %>% 
  mutate(name = "US", fips = "00")

```

## Let's join across geographies and split across metrics
```{r}
all_debt_share_amount <- bind_rows(county_share_amount, state_share_amount, nation_share_amount)

all_share_with_debt_all <- all_debt_share_amount %>% 
  select(fips, name, year, contains("share"), -contains("majority")) %>% 
  pivot_wider(names_from = year, values_from = share_with_medical_debt_in_collections) %>% 
  mutate(metric = "share_debt_all")

all_share_with_debt_poc <- all_debt_share_amount %>% 
  select(fips, name, year, share_with_medical_debt_in_collections_majority_of_color) %>% 
  pivot_wider(names_from = year, values_from = contains("share")) %>% 
  mutate(metric = "share_debt_poc")

all_share_with_debt_white <- all_debt_share_amount %>% 
  select(fips, name, year, share_with_medical_debt_in_collections_majority_white) %>% 
  pivot_wider(names_from = year, values_from = contains("share")) %>% 
  mutate(metric = "share_debt_white")

# ------------------------------

all_amount_of_debt_all <- all_debt_share_amount %>% 
  select(fips, name, year, contains("median"), -contains("majority")) %>% 
  pivot_wider(names_from = year, values_from = contains("median")) %>% 
  mutate(metric = "amount_debt_all")

all_amount_of_debt_poc <- all_debt_share_amount %>% 
  select(fips, name, year, median_medical_debt_in_collections_in_2023_majority_of_color) %>% 
  pivot_wider(names_from = year, values_from = contains("median")) %>% 
  mutate(metric = "amount_debt_poc")

all_amount_of_debt_white <- all_debt_share_amount %>% 
  select(fips, name, year, median_medical_debt_in_collections_in_2023_majority_white) %>% 
  pivot_wider(names_from = year, values_from = contains("median")) %>% 
  mutate(metric = "amount_debt_white")

```

# Import all files for metric charts

## ACS data
```{r}
# county ACS data
county_acs <- read_xlsx('./data-in/acs_county_2023.xlsx') %>% 
  clean_names() %>% 
  select(-county_nm, -state_abbreviation) %>% 
  rename(fips = county_fips)

# state ACS data
state_acs <- read_xlsx('./data-in/acs_state_2023.xlsx') %>% 
  clean_names() %>% 
  select(-state_abbreviation) %>% 
  rename(fips = state_fip)

# us ACS data
nation_acs <- read_xlsx('./data-in/acs_national_2023.xlsx') %>% 
  clean_names() %>% 
  select(-usa) %>% 
  mutate(fips = "00")

all_acs <- bind_rows(county_acs, state_acs, nation_acs) %>% 
  mutate(year = as.character(year))

all_no_insurance <- all_acs %>% 
  select(fips, year, share_of_the_population_with_no_health_insurance_coverage) %>% 
  mutate(share_of_the_population_with_no_health_insurance_coverage = as.character(share_of_the_population_with_no_health_insurance_coverage)) %>% 
  pivot_wider(names_from = year, values_from = share_of_the_population_with_no_health_insurance_coverage) %>% 
  mutate(metric = "uninsured")

all_income <- all_acs %>% 
  select(fips, year, average_household_income_in_2023_dollars) %>% 
  mutate(average_household_income_in_2023_dollars = as.character(average_household_income_in_2023_dollars)) %>% 
  pivot_wider(names_from = year, values_from = average_household_income_in_2023_dollars) %>% 
  mutate(metric = "hh_income")

all_disability <- all_acs %>% 
  select(fips, year, share_of_non_elderly_adults_with_a_reported_disability) %>% 
  mutate(share_of_non_elderly_adults_with_a_reported_disability = as.character(share_of_non_elderly_adults_with_a_reported_disability)) %>% 
  pivot_wider(names_from = year, values_from = share_of_non_elderly_adults_with_a_reported_disability) %>% 
  mutate(metric = "share_disabled")

```

## Merger data
### Notes
These files coming from the research team remove columns for regions where no merges or closures occured. That means that in joining these files, not every region is guaranteed to have data for this metric, and I can't assume it'll be present. Therefore, I'll check if data was found for a certain metric for a certain region, and if the metric is "closes_merges" I can fill with zeroes -- but if the metric is not this one, I'll still throw an error. 

```{r}
# county merger data
county_merger <- read_xlsx('./data-in/close_merge_sum_county.xlsx') %>% 
  clean_names() %>% 
  rename(fips = county_fips)

# state merger data
state_merger <- read_xlsx('./data-in/close_merge_sum_state.xlsx') %>% 
  clean_names() %>% 
  select(-state_name) %>% 
  rename(fips = state_fip)

# us merger data
nation_merger <- read_xlsx('./data-in/close_merge_sum_nat.xlsx') %>% 
  clean_names() %>% 
  select(-national) %>% 
  mutate(fips = "00")

all_merger <- bind_rows(county_merger, state_merger, nation_merger) %>% 
  mutate(number_of_hospital_closures_mergers = as.character(number_of_hospital_closures_mergers)) %>% 
  pivot_wider(names_from = year, values_from = number_of_hospital_closures_mergers) %>% 
  mutate(metric = "closes_merges")


```

## Hospital density data
```{r}
# county hospital density data
county_density <- read_xls('./data-in/HHI by year and county.xls') %>% 
  clean_names() %>% 
  rename(fips = county, name = countyname)

# state hospital density data
state_density <- read_xls('./data-in/HHI by year and state.xls') %>% 
  clean_names() %>% 
  rename(fips = state, name = stab)

# us hospital density data
nation_density <- read_xls('./data-in/HHI by year and national.xls') %>% 
  clean_names() %>% 
  rename(hhi_w = hhi_unweight) %>% 
  select(-us) %>% 
  mutate(fips = "00", name = "US")

all_density <- bind_rows(county_density, state_density, nation_density) %>% 
  mutate(year = as.character(year), hhi_w = as.character(hhi_w)) %>% 
  pivot_wider(names_from = year, values_from = hhi_w) %>% 
  mutate(metric = "hospital_density")
```

## make a tidy little name table for everyone

```{r}
st_crosswalk <- tibble(state = state.name) %>%
   bind_cols(tibble(abb = state.abb)) %>% 
   bind_rows(tibble(state = "District of Columbia", abb = "DC"))
  
names <- all_debt_share_amount %>% 
  select(fips, region_lsad = name, state_abreviation) %>% 
  mutate(state_abreviation = ifelse(nchar(region_lsad) == 2, region_lsad, state_abreviation)) %>% 
  distinct() %>% 
  left_join(st_crosswalk, by = c("state_abreviation" = "abb")) %>% 
  mutate(full_name = ifelse(region_lsad == "US", "United States of America",
                            ifelse(nchar(region_lsad) == 2, state, 
                                   paste0(region_lsad,", ",state)))) %>% 
  select(fips, full_name)

```

# Now we're going to take all these metric tables and join em up

```{r}
one_big_data_by_geo <- bind_rows(all_disability, all_income, all_no_insurance,
                                 all_amount_of_debt_all, all_amount_of_debt_poc, all_amount_of_debt_white,
                                 all_share_with_debt_all, all_share_with_debt_poc, all_share_with_debt_white,
                                 all_density, all_merger) %>% 
  select(-name) %>% 
  left_join(names, by = c("fips")) %>% 
  select(fips, full_name, metric, everything())

write_csv(one_big_data_by_geo, "./data-out/one_big_data_by_geo.csv") # 5.4 MB
write(toJSON(one_big_data_by_geo),"./data-out/one_big_data_by_geo.json") # 9.7 MB

```









# For initial EDA in Datawrapper

```{r}
# meddebt_states <- read_excel("data-in/all_state_export.xlsx") %>% 
#   clean_names() 
# 
# median_meddebt <- meddebt_states%>% 
#   select(-share_with_medical_debt_in_collections, -state_fip) %>% 
#   pivot_wider(names_from = state_abreviation, values_from = median_medical_debt_in_collections_in_2023) %>% 
#   write_csv("data-out/state_median_meddebt.csv")
# 
# share_meddebt <- meddebt_states%>% 
#   select(-median_medical_debt_in_collections_in_2023, -state_fip) %>% 
#   mutate(share_with_medical_debt_in_collections = round(100*as.double(share_with_medical_debt_in_collections),1)) %>% 
#   pivot_wider(names_from = state_abreviation, values_from = share_with_medical_debt_in_collections) %>% 
#   write_csv("data-out/state_share_meddebt.csv")
# 
# glimpse(meddebt_states)
```

```{r}
# meddebt_states_hhi <- read_excel("./data-in/HHI by year and state.xls") %>% 
#   clean_names() 
# 
# hhi_by_state <- meddebt_states_hhi%>% 
#   mutate(hhi_w_clean = round(hhi_w)) %>% 
#   select(-state, -hhi_w) %>% 
#   pivot_wider(names_from = stab, values_from = hhi_w_clean) %>% 
#   write_csv("./data-out/hhi_state.csv")
# 
# 
# glimpse(meddebt_states)


```